---
title: "Analysis of Storm data for better allocation of resources for storm events"
output: 
  html_document:
    keep_md: true
---
```{r include=FALSE}
knitr::opts_chunk$set(echo=TRUE,warning = FALSE)
```

## Abstract

Various Events such as Typhoon, Extreme heat, Floods , Tsunami etc can cause huge losses of life and property. It is hard to prevent these events but we can be prepared and show extra care when we know that the event can cause a large scale damage in terms of human life and economy.

The dataset provided by US National Oceanic and Atmospheric Administraion's (NOAA) storm database consists of records of various events across the years and an estimate to the number of deaths, injuries, property damage estimate and crop damage estimate. This research analyses this data and provides an insight to the events that cause the maximum damage so that allocation of resources during an event might be optimised.

## Reading the Data

If data has not been downloaded then the data can be downloaded from the following link:   
[Storm dataset (47 MB)](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)   

The file has to be unzipped and stored in the same directory

```{r,cache=TRUE}
# To check if file exists in the directory uncomment the below line
# file.exists("repdata_data_StormData.csv")

# Reading the data
data <- read.csv("repdata_data_StormData.csv")

```

Displaying the data

```{r}
head(data)
```

## Data Processing

### Loading Required Libraries for processing

Some packages are required to be installed. Then they have to be imported with the following command.

```{r,warning=FALSE}

# importing data.table library
# If not installed uncomment the below line first
# install.packages("data.table")
library(data.table)

# importing dplyr library
# If not installed uncomment the below line first
# install.packages("dplyr")
library(dplyr)

```

### Subsetting the data for events and fatalities

A new dataframe is created by grouping events(EVTYPE) and then applying sum function on fatalities as follows:   
```{r}
# Grouping on the basis of evtype and applying sum function
fatalities_evtype <- aggregate(FATALITIES~EVTYPE,data,sum)

# Sorting on the basis of fatalities
fatalities_evtype <- arrange(fatalities_evtype,desc(FATALITIES))

```

### Subsettting the data for events and injuries

A new dataframe is created by grouping events(EVTYPE) and then applying sum function on injuries as follows:   
```{r}
# Grouping on the basis of evtype and applying sum function
injuries_evtype <- aggregate(INJURIES~EVTYPE,data,sum)

# Sorting on basis of injuries(descending)
injuries_evtype <- arrange(injuries_evtype,desc(INJURIES))
```

### Calculation of Property damage and Crop Damage

Individual Property damage and Crop damage is calculated based on the abbrevation given in PROPDMGEXP and CROPDMG column

The description of the abbrevation can be found by clicking [here](https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html)

Abbrevations :

* H,h = Hundreds   = x 100
* K,k = Thousands  = x 1,000
* M,m = Millions   = x 1,000,000
* B,b = Billion    = x 1,000,000,000
* (+) = NA         = x 1
* (-) = NA         = x 0
*  ?  = NA         = x 0
*  "" = NA         = x 0

The following code calculates total damage
```{r}
# Creating new dataframe with required columns
economy <- data.frame(data$EVTYPE,data$PROPDMG,data$PROPDMGEXP,data$CROPDMG,data$CROPDMGEXP)

# Assigning names to columns
names(economy) <- c("EVTYPE","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")

# Getting the abbrevations
Symbol <- sort(unique(as.character(economy$PROPDMGEXP)))

# Creating a vector with values to abbrevations
Multiplier <- c(0,0,0,1,10,10,10,10,10,10,10,10,10,10^9,10^2,10^2,10^3,10^6,10^6)

# Creating a data frame with abbrevations and their values
c.Multiplier <- data.frame(Symbol, Multiplier)

# Matching PROPDMGEXP abbrevation with the one in Symbol list
economy$Prop.Multiplier <- c.Multiplier$Multiplier[match(economy$PROPDMGEXP, c.Multiplier$Symbol)]

# Matching CROPDMGEXP abbrevation with the one in Symbol list
economy$Crop.Multiplier <- c.Multiplier$Multiplier[match(economy$CROPDMGEXP, c.Multiplier$Symbol)]

# Calculating Total Economic damage
economy <- economy %>% mutate(PROPDMG = PROPDMG*Prop.Multiplier) %>% mutate(CROPDMG = CROPDMG*Crop.Multiplier) %>% mutate(Total_damage = PROPDMG+CROPDMG)

# Grouping on the basis of Event and applying sum function
economy <- aggregate(Total_damage~EVTYPE,economy,sum)

# Sorting on basis of Total_damage(Descending)
economy <- arrange(economy,desc(Total_damage))

# Displaying few values
head(economy)


```

## Results

### Events with Fatalities

Top 20 highest fatality events are shown.

```{r Fatalities,fig.height=8,fig.width=12}
# importing required library
library(ggplot2)

# Plotting graph
# Defining x,y and fill bars on the basis of Event Type
g <- ggplot(fatalities_evtype[1:20,],
            aes(reorder(EVTYPE,-FATALITIES),
                FATALITIES,
                fill=EVTYPE))

# Plotting a bargraph
g+geom_bar(stat="identity")+
  
  # Defining labels
  labs(x="Events",
       title = "Top 20 Events with Fatalities",
       fill="Event")+
  
  # Defining theme
  theme(axis.text.x =element_text(angle = 90,hjust = 1))
```

Tornado caused the maximum deaths with `r fatalities_evtype$FATALITIES[1]` deaths.

### Events with Injuries

Top 20 highest injury causing events are shown.

```{r Injuries,fig.height=8,fig.width=12}
# Plotting graph
# Defining x,y and fill bars on the basis of Event Type
g <- ggplot(injuries_evtype[1:20,],
            aes(reorder(EVTYPE,-INJURIES),
                INJURIES,
                fill=EVTYPE))

# Plotting a bargraph
g+geom_bar(stat="identity")+
  
  # Defining labels
  labs(x="Events",
       title = "Top 20 Events with Injuries",
       fill="Event")+
  
  # Defining theme
  theme(axis.text.x =element_text(angle = 90,hjust = 1))
```

Tornado caused the maximum injuries with `r injuries_evtype$INJURIES[1]` injuries.

### Events with Maximum Economic Damage

Top 20 highest ecomomy impact events are shown.

```{r Economic_damage,fig.height=8,fig.width=12}
# Plotting graph
# Defining x,y and fill bars on the basis of Event Type
g <- ggplot(economy[1:20,],
            aes(reorder(EVTYPE,-Total_damage),
                Total_damage,
                fill=EVTYPE))

# Plotting a bargraph
g+geom_bar(stat="identity")+
  
  # Defining labels
  labs(x="Events",
       y="Total Impact (US$)",
       title = "Top 20 Events with Maximum Economic damage",
       fill="Event")+
  
  # Defining theme
  theme(axis.text.x =element_text(angle = 90,hjust = 1))
```

Flood caused the maximum economic impact with overall loss of `r economy$Total_damage[1]` dollars.